---
title: "MB5370 Module 04. Workshop 1 - Introduction"
author: "Ben Cresswell"
output: html_document
date: "2024-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Install and load tidyverse packages
```{r}

library("tidyverse")
```


# Load the data (mpg is built into ggplot2)
```{r}
data(mpg)
```


# Quick data checks
```{r}
head(mpg)
glimpse(mpg)
summary(mpg)
```


# Create first plot
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?



# Understanding grammar of graphics
What happens if we just try to run ggplot on its own?
```{r}
ggplot()
# Or with just the data specified?
ggplot(data = mpg)
```

We need to map some aesthetics!!

# When you’re creating a plot, you essentially need two attributes of a plot: a geom and aesthetics.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
```


# Change point colour by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) #gives random colors to classes
```

# Change point size by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class)) #gives ordered size to classes
```
Note the warning!!!


# Change transparency (alpha) by class
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class)) #gives ordered alpha(shades) to classes
```
Another warning!!
Question: When would using alpha (or size be appropriate??) ######
#when dealing with overlapping, density size, or adding emphasis

# Change point shape by class:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class)) #gives different shape
```


# Make all points blue
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") #made the entire class blue
```


# Troubleshooting 






# Faceting

## facet_wrap
Split the plot out by car type (class)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = , ncol = 7) #ditinguishes all the class variables in different graphs
```
#the plus sign should never be at the start of the code but at the end of the first line, or else it wont run.

## facet_grid
A separate facet for each combination of drive-type (e.g. 4WD) * number of cylinders
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl) #ask chatgpt? whats the diff between grid and wrap #ecologically, used to compare the relationship between displacement and highway.
```
Note that there are no occurrences of 5 cylinder 4WDs OR RWD vehicles!


## facet_grid by just row (or column)
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(cyl ~ .)
  #facet_grid(. ~ cyl) # Alternatively #removed the vertical gird as we put dot instead of hwy

```

# Exercise:
Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? 

```{r}
#gives the number of rows and columns to be added to the facet_wrap.scales = "fixed",
  #shrink = TRUE,
  #labeller = "label_value",
  #as.table = TRUE,
  #switch = deprecated(),
  #drop = TRUE,
  #dir = "h",
  #strip.position = "top",
  #axes = "margins",
  #axis.labels = "all"

```


# Lines
We already used points:
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displx, y = hwy))
```


However, ggplot2 can use a variety of geom objects to represent the data. Here, we might want to use bar plots, line charts, boxplots and so on. Well we can handle this issue in ggplot directly using a different geom to plot the same data. Here, instead of plotting points, we will use a smooth line. 

# To display the same data as a smooth line fit through the points use geom_smooth().
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```
### Question: whey don't we use geom_line() here? What would that look like? ### use geom_line when you have continuous variables(means the variables that will always keep increasing).if we ran it in categorical instead of continuous it would give zig-zag lines


So let’s recap. A geom is an object that your plot uses to represent the data. To change the geom type in your plot, simply change the geom function that you add to your plot template. Sometimes you may want to try a few things out, in which case you could use comments to help you remember what worked and what didn’t. 

## Using comments (#)
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy)) + # points horrible 
  geom_smooth(mapping = aes(x = displ, y = hwy)) # try smooth line
```

Question: how does R work out where to plot the line??? Can use the chunk output to inform us. Can also use the help menu.

# Changing linetype
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) # Can also use "lty = " for the same result #different line types for the objects, classes that you are trying to distinguish
```


# Grouping
```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = class)) #groups the lines according to the class, or object.
```
 
# Change line colour based on drv value
```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = class),
    show.legend = FALSE,
  ) #color will add color to each class.
```

# Multiple geoms
We already did this one way
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
#this code will give point and line graph#but this is not ideal in programming sense as the x and y variables are repetative so if we change at one place will have to change them at many places.
```

A better way...
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
#put x and y in ggplot.
```


# Can still manipulate each geom/layer separately:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth() #mapping correct layer(lines) of code 
```

Now we have different colours to indicate the raw data (geom_point) and one colour for the overall relationship (geom_smooth). Neat!


```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)

```
Question: what does the "se = FALSE" do?


# Exercise:
1.  What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
#do this exercise 
```{r}

```


3.  Will these two graphs look different? Why/why not? #its actually the same code
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```


# Transformations and stats

## Load data
```{r}
data("diamonds")
glimpse(diamonds)
```

# Bar plot
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

# Overriding defaults
Make some new data
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
```


```{r}
ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
Question: Why did we specify now specify a y axis?

# What if we want to know the proportion rather than the absolute frequency?
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```

Question: does anyone get the warning that's in the workbook?
Warning: `stat(prop)` was deprecated in ggplot2 3.4.0.
If so, can use # geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), group = 1))

# Plotting statistical details
```{r}
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```


# Aesthetic adjustments adjustments
Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like colour or fill to change aspects of bar colours. We already did this once, but there are multiple options available to you, including changing the fill our outline colours.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```
Question: Does anyone notice anything different in the colour scheme of this plot?
(Hint: It's in the viridis colour palette (colour blind friendly), but why is it different from the colour palette we used earlier?)
Check out the difference:
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) 

ggplot(data = diamonds, mapping = aes(x = color, y = depth)) + 
  geom_point(mapping = aes(color = cut)) 
```


# Filling by a variable
Now try using these aesthetics to colour by another variable like clarity. Notice how the stacking is done automatically. This is done behind the scenes with a position argument.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```


# To alter transparency (alpha)
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
```


# To color the bar outlines with no fill color
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```


# Position adjustments
position = "fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```


# position = "dodge" 
Places overlapping objects directly beside one another.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

# Jittering
position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data = mpg) + 
  #geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
  geom_point(mapping = aes(x = displ, y = hwy))
```


#Labels
Labs() function adds label which help you communicate your key findings. Good labels communicate key findings better and also comply with formatting requirements of scientific journals.

```{r}
ggplot(mpg, aes(displ, hwy)) + #replacked x & y with displ and hwy
  geom_point(aes(color = class)) +
  geom_smooth(se.e = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")

```

subtitles usually in smaller font, and captions at the bottom right of the plot, often used to describe the source of the data.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  ) #can use to cite a graph if you got it from somewhere
```

Can also use labs() to replace axis labels and legend titles

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )

```
you can also make mathematical notations and superscripts with labels.

#Annotations
what if we want to add text to the plot itself? we can use geom_text() for that which works the same as geom_point, where insted of a geometric output it will give a label. Also now we will filter data and find out a specific value that we need from the data like ' the most fuel efficient car' and then lable it.
```{r}
best_in_class <- mpg %>% #why given another variable? to avoid confusion
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)
#can use nudge function to remove the overlapping of the variables

```


#Scales
changing scales customises plots and imrpoves communication of your results. You can do so by offering values to the scale parameters by adding numbers in the appropriate scale arguments.you will have to provide a character vector for putting limits to your scale(eg. c(1,12)). scale_x_continuous has plenty other arguements, so dont forget to explicitly mention your arguement (limits = c(1,12)).

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete() #didnt change anything, maybe because no arguement was applied
```

#Axis Ticks

changes ticks on your axis. breaks = controls the position of the ticks and we can also add labels as arguements to them.

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5)) #breaks the y axis by 5 from 15-40
```
Using labels now

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)#I will have to create a character variable for adding words in the axis
```
#Legends and Color Schemes

theme() to control the position of legend however you want.

```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") # the default and legend.position = "none" to make legend go away
```

#Replacing a scale

2 types of scales : continuous position and colour scales. transformation of your variables.

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10() #comparison of carat to price
```
geom_bin2d rpovides rectangular shaped graph

color scales
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
```

Rebundant shape mapping if only few colors, so can also be interpretable in black and white.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")
```
The ColorBrewer scales are documented online at http://colorbrewer2.org/. There are a range of different colour schemes you can import such as this, so read more about them in the ggplot2 cookbook: http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/ 

scale_color_manual to set colors yourself.

```{r}
library(ggplot2)
library(dplyr)
presidential %>% #these "%%" signs take the first step object and pass it on to the next function. presidential is a ggplot function.
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue")
```
using bens color package
```{r}
install.packages('viridis')
install.packages('hexbin')
library(viridis)
library(hexbin)
df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() + #I like the geom_hex!
  viridis::scale_fill_viridis() +
  coord_fixed()

```

#themes
customising themes of your plot, allow you to customise some or all of your non-data elements of your plot with a theme. ggplot has 8 themes by default Many more are included in add-on packages like ggthemes (https://github.com/jrnold/ggthemes), by Jeffrey Arnold.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()

```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()


```
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()

```
you can develop your own theme by setting some or all of your arguements to the theme yourself

```{r}
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial"))

#I still have to know how this is working

```
#saving and exporting your data
ggsave() helps you save your most recent plot to your working directory.

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
#have to try width and height function 
#> Saving 7 x 4.32 in image
```

